---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

To load the data:
```{r}
activityData <- read.csv("activity.csv")
summary(activityData)
head(activityData)
```
I decided to convert the "date" column into the Date format.
```{r}
activityData$date <- as.Date(activityData$date)
```
Now the data is ready to use.
## What is mean total number of steps taken per day?
I split the data by day, then summed the steps on each day.
```{r, results='asis'}
a <- split(activityData, activityData$date)
sum <-sapply(a, function(x) {sum(x[, c("steps")], na.rm = TRUE)})
```
```{r, echo=FALSE}
hist(sum, c(10), main ="Total number of steps per day" )
```
```{r, echo=FALSE}
meanstep<-mean(sum, na.rm = TRUE)
medianstep<-median(sum, na.rm = TRUE)
```
The mean number of steps a day is `r meanstep` and the median is `r medianstep`.

## What is the average daily activity pattern?
To plot the daily activity pattern I split the data by interval then found the mean for each interval.
```{r}
interval<-unique(activityData$interval)
b<- split(activityData,interval)
pattern <- sapply(b, function(x) {mean(x[, c("steps")], na.rm = TRUE)})
```
```{r, echo=FALSE}
plot(interval,pattern, type = "l", main ="Average daily activity pattern",xlab ="Interval",ylab = "Average number of steps")
```
```{r, echo=FALSE}
d<-data.frame(interval,pattern)
maxd <-d[d$pattern == max(d$pattern),]
```
The 5 minute interval is interval `r maxd[1]` with `r maxd[2]` steps.

## Imputing missing values

```{r,echo=FALSE}
numna<-nrow(activityData[is.na(activityData$steps),])
```
The number of rows containing NAs is `r numna`.
I decided to fill in the missing values with the interval average.
```{r}
completeActivityData<-activityData
completeActivityData$steps <- ifelse(is.na(completeActivityData$steps), d$pattern[match(completeActivityData$interval, d$interval)], completeActivityData$steps)
```
```{r, echo=FALSE}
newnumna<-nrow(completeActivityData[is.na(completeActivityData$steps),])
```
Now the number of rows containing NAs is `r newnumna`.
```{r , results='asis'}
m <- split(completeActivityData, completeActivityData$date)
newsum <-sapply(m, function(x) {sum(x[, c("steps")], na.rm = TRUE)})
```
```{r, echo=FALSE}
hist(newsum, c(10), main ="Mean steps per day with missing values filled in")
```
```{r, echo=FALSE}
newmeanstep<-mean(newsum, na.rm = TRUE)
newmedianstep<-median(newsum, na.rm = TRUE)
```
The mean number of steps a day is `r format(newmeanstep,scientific = FALSE)` and the median is `r format(newmedianstep, scientific = FALSE)`. Compare that to the old mean:  `r round(meanstep, 2)` and median: `r medianstep`. The new results are higher.

## Are there differences in activity patterns between weekdays and weekends?
I created a new column "weekday" that contains TRUE on a weekday and FALSE on the weekend
```{r}
completeActivityData$weekday<-weekdays(completeActivityData$date)
completeActivityData$weekday<-ifelse(completeActivityData$weekday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), TRUE, FALSE)
```
I then split my data into two data frames and used the same pattern formulas I used earlier to both.
```{r}
weekdaydata<-completeActivityData[completeActivityData$weekday,]
weekenddata<-completeActivityData[!completeActivityData$weekday,]
weekdaysplit<- split(weekdaydata,weekdaydata$interval)
weekendsplit<- split(weekenddata,weekenddata$interval)
weekdaypattern <- sapply(weekdaysplit, function(x) {mean(x[, c("steps")], na.rm = TRUE)})
weekendpattern <- sapply(weekendsplit, function(x) {mean(x[, c("steps")], na.rm = TRUE)})
```
```{r, echo=FALSE}

par(mfrow=c(2,1))
par(mar=c(0,0,0,0),oma = c(4, 4, 4, 4))
plot( interval, weekdaypattern, type="l", col="red")
plot( interval, weekendpattern, type="l", col="green")
mtext("Average daily activity pattern",side=3,outer=TRUE,padj=-2)
mtext("Interval",side=1,outer=TRUE,padj=3)
mtext("Average number of steps",side=2,outer=TRUE,padj=-4)
mtext("weekend               weekday",side=4,outer=TRUE,padj=1)
```





